\chapter{Conclusion}
\label{ch:conclusion}

Motivated by the need to improve real-time situational awareness of power system operators and the emergence of the PMU technology, this thesis sets out to develop advanced data analytic methods for the detection and identification of transmission lines in real time. 

In Chapter \ref{ch:detection_using_approximate_dynamics}, a novel line outage detection scheme is proposed based on the power flow model and generalized likelihood ratio testing. The power flow model is used as the basis for predicting post-outage angle deviations. The control chart constructed based on the GLR procedure is able to detect any outages quickly while controlling for the false alarm rate. Notably, the proposed method can capture system dynamics since it retains the time-variant and nonlinear nature of the power system. Extensive simulation study suggests that the method performs well for outage detection where most outages can be detected with less than one second of delay. However, longer delays are observed for outage lines which have no nearby PMU sensors, prompting the need for research into more robust detection scheme. 

Chapter \ref{ch:detection_using_generator_dynamics} extends the research in the previous chapter by proposing a unified detection framework where both generator dynamics and load bus dynamics can be monitored. The unified framework consists of a particle filter-based nonlinear state estimator and a MEWMA-based control chart. Through simulation study using the IEEE 39-bus system, it is shown that the inclusion of generator dynamics makes detection faster and more robust to a priori unknown outage locations. In particular, 80\% of the simulated outages can be detected and most of them by 0.2 seconds after the event. 

Although encouraging results are seen from the previous two chapters, they do not address another important and practical question regarding line outage awareness - identification. Assuming an outage detection alarm has been raised, Chapter \ref{ch:identification} proposes a novel method to accurately identify a priori unknown number of outage lines using a limited PMU sensors. This work draws inspiration from sensitivity analysis of the AC power flow model and advances in underdetermined sparse regression methods. The use of lasso formulation and LARS algorithm, in particular, overcomes the inherent combinatorial challenge of the line identification problem. Compared to the state-of-the-art work in multiple-line outage identification, the proposed method is able to achieve over 90\% and 80\% accuracy for single- and double-line outage identification. 

The guiding principal to this thesis' approach to address the outage detection and identification problem is a deep integration between statistical monitoring and diagnostic methods and power system physical modeling. Through three chapters of work, it is demonstrated that the AC power flow model, although nonlinear and more complex than the DC version commonly used, is able to provide significant advantage in both detection and identification. Still, the performance is achieved by combining the physical model with careful application of methods outside the traditional domain of power system research, e.g., from the field of statistical quality control and data mining. As PMU technology takes a central role in the modernization drive of power grids across the world, real-time outage detection and identification methods developed in this thesis have the potential to contribute to a more reliable and resilient power system. 

There are three worthwhile future research directions stemming from the work in this thesis: (1) The first concerns the optimal placement of limited PMUs in a network that maximizes identification accuracy. As seen in Chapter \ref{ch:identification}, the placement of PMUs influence the diagnosability of the system, suggesting the potential reward of a carefully considered placement. In general, the placement of sensors is usually formulated as a mixed-integer programming problem that is difficult to solve exactly. Also, the impact of the number of available PMUs on the identification performance could be studied. (2) The second area of future research is active diagnostics for indistinguishable line outages. The problem of ambiguous outage lines have been observed in previous research and this thesis. Until now, most research effort have focused on passively analyzing power system data in order to ascertain the location of outage lines. However for the group of line outages which might be difficult to distinguish from one another, proactive identification methods might be needed. For example, targeted system inputs can be employed to specifically find out the most probably outage line among a group of highly similar candidate lines. (3) Lastly, more research can be done in the area of optimal post-disruption recovery for power systems. With contingencies like line outage detected and identified, system operators need to coordinate recovery actions to minimize the potential impact while considering various constraints. Also, uncertainties in resources available and repair duration of line outages need to be factored in when designing a recovery strategy. The ability to recover efficiently following a disruption is a critical aspect of resilient power systems. 


