\chapter{Conclusion}
\label{ch:conclusion}

Motivated by the need to improve real-time situational awareness of power system operators and the emergence of the PMU technology, this thesis sets out to develop advanced data analytic methods to detect and identify transmission line outages in real-time.


\section{Contributions}
In Chapter \ref{ch:detection_using_approximate_dynamics}, a novel line outage detection scheme is proposed based on the power flow model and generalized likelihood ratio testing. The power flow model is used as the basis for predicting post-outage angle deviations. The control chart constructed based on the GLR procedure can detect any outages quickly while controlling for the false alarm rate. Notably, the proposed method can capture system dynamics since it retains the time-variant and nonlinear nature of the power system. An extensive simulation study suggests that the method performs well for outage detection, where most outages can be detected with less than one second of delay. However, longer delays are observed for outage lines with no nearby PMU sensors, prompting the need for research into a more robust detection scheme.

Chapter \ref{ch:detection_using_generator_dynamics} extends the research in the previous chapter by proposing a unified detection framework that monitors both generator dynamics and load bus dynamics. The unified framework consists of a particle filter-based nonlinear state estimator and a MEWMA-based control chart. The simulation study using the IEEE 39-bus system shows that the inclusion of generator dynamics makes detection faster and more robust to a priori unknown outage locations. In particular, 80\% of the simulated outages can be detected, and most of them are detected by 0.2 seconds after the event. 

Although encouraging results are seen from the previous two chapters, they do not address another critical and practical question regarding line outage awareness - identification. Assuming an outage detection alarm has been raised, Chapter \ref{ch:identification} proposes a novel method to accurately identify a priori unknown number of outage lines using limited PMU sensors. This work draws inspiration from sensitivity analysis of the AC power flow model and advances in underdetermined sparse regression methods. The use of lasso formulation and the LARS algorithm, in particular, overcomes the inherent combinatorial challenge of the line identification problem. Compared to the state-of-the-art in multiple-line outage identification, the proposed method achieves over 90\% and 80\% accuracy for single- and double-line outage identification. 

The guiding principle to this thesis' approach to addressing the outage detection and identification problem is deep integration between statistical monitoring and diagnostic methods and power system physical modeling. Through three chapters of work, it is demonstrated that the AC power flow model, although nonlinear and more complex than the DC version commonly used, can provide a significant advantage in both detection and identification. Still, the performance is achieved by combining the physical model with careful application of methods outside the traditional domain of power system research, e.g., statistical quality control and data mining. As PMU technology takes a central role in the modernization drive of power grids across the world, real-time outage detection and identification methods developed in this thesis can contribute to a more reliable and resilient power system. 

\section{Limitations}
There are several limitations in the studies conducted in this thesis. Although important, the optimal placement of PMU sensors is not investigated in depth in all of our studies. Also, given the size of a power system, the minimum number of PMUs required to achieve a certain level of detection and identification percentage is not investigated. On the other hand, the emerging issue of PMU data quality is also not addressed in this thesis. In all three studies, the PMU data is assumed to be corrupted with only normal sensor errors and random load fluctuations. Bad PMU data due to sensor error or maliciously injected false data could impact the effectiveness of the detection and identification schemes proposed in this thesis. Also, the concept of minimal diagnosable cluster of lines proposed in the third study is an approximation of the distinctiveness between outages. They are limited to linear correlations between a pair of transmission lines. More general definition of similarity and diagnosability between an arbitrary number of lines could be proposed. 

\section{Future Research Directions}
Three future research directions stemming from this thesis are worth pursuing: (1) The first concerns the optimal placement of limited PMUs in a network that maximizes detection speed, detection rate, and identification accuracy. Firstly, although increasing the number of PMUs leads, in general, to an increased performance in terms of detection speed, rate, and identification accuracy, the performance gain seems to diminish quickly as more PMUs become available. One could wish to study the minimum number of PMUs needed to achieve a certain level of detection speed, rate, or identification accuracy. Secondly, for fast outage detection without using generator state dynamics, e.g., the proposed method in Chapter \ref{ch:detection_using_approximate_dynamics}, a scattered and spread out set of PMU locations tend to achieve better detection speed compared to a set of concentrated locations. On the other hand, for the proposed method in Chapter \ref{ch:detection_using_generator_dynamics}, it requires PMUs to be installed on both generator and load buses. PMU location optimization in this case has to accommodate this requirement. For example, the optimization problem might need to be divided into two subproblems, one for generator bus PMUs and one for load bus PMUs. Lastly, as seen in Chapter \ref{ch:identification}, the placement of PMUs influences the diagnosability of the system, suggesting the potential reward of a carefully considered placement. For example, PMUs can be located in a way such that the orthogonality of the columns in the phase angle signature matrix are maximized. (2) The second area of future research is active diagnostics for indistinguishable line outages. The problem of ambiguous outage lines has been observed in previous research and this thesis. Until now, most research efforts have focused on passively analyzing power system data to ascertain the location of outage lines. However, for the group of line outages that might be difficult to distinguish from one another, proactive identification methods might be needed. For example, targeted system inputs can be employed to specifically find out the most probably outage line among a group of highly similar candidate lines. (3) Lastly, more research can be done in the area of optimal post-disruption recovery for power systems. With contingencies like line outages detected and identified, system operators need to coordinate recovery actions to minimize the potential impact while considering various constraints. Also, uncertainties in resources available and repair duration of line outages need to be factored in when designing a recovery strategy. The ability to recover efficiently following a disruption is a critical aspect of resilient power systems. 


