\begin{acknowledgments}

I would like to acknowledge the contribution of many people to the development of this thesis and my Ph.D. journey. Without them, this work would not have been possible, and I would not be the same. 

I am most grateful to my Ph.D. advisor, Prof. Chen Nan, who has greatly influenced my professional and personal development over the last four years. His insistence on integrating domain knowledge and meticulous approach to contribution justification shaped my thinking of each research problem. He has provided a liberal and safe environment where I could learn continuously, form opinions, and discover insights every day. He has encouraged me on numerous occasions to think differently, dare answer a more difficult question, and step into new territories. The guidance from him has helped me become a better researcher and a better person. For that, I will be forever grateful.  

I would also like to thank my collaborator, Prof. Zhai Chao. He provided much assistance and knowledge throughout the development of this thesis. In particular, his knowledge of power systems helped me tremendously during the early phase of my Ph.D. journey. I am fortunate to be among a group of motivated and supportive group mates. They are Shi Yuchen, Zhu Jun, Wang Rui, and Liang Hongde. They have always generously offered me words of encouragement and ideas from different perspectives. 

I thank the Singapore-ETH Centre and Singapore's National Research Foundation for the financial support of my Ph.D. study. I also benefited significantly by working with colleagues from the Future Resilient Systems program at the SEC. They provided a comfortable and stimulating research environment for me. 

Finally, I would like to thank my parents, Zhou Yonghong and Yang Zhenglu, and my girlfriend, Chen Linye, for their unwavering love and support. This thesis is dedicated to you all. 


\end{acknowledgments}
